---
title: "Jade_statsreport"
output: html_document
date: "2023-11-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Install and load necessary packages
install.packages("readxl")
library(readxl)
install.packages("dplyr")
library(dplyr)
install.packages("zoo")
library(zoo)
install.packages("TTR")
library(TTR)
install.packages("forecast")
library(forecast)
install.packages("ggplot2")
library(ggplot2)
install.packages("Metrics")
library(Metrics)
install.packages("tseries")
library(tseries)
```

```{r}
# Read and check excel data
c11 <- read_excel("/Users/jadeoh/RStudio/eco_time_series/asobara_only_0refuge/c11.xlsx")
head(c11)
any(is.na(c11$TotalFlies))
any(is.na(c11$TotalPara))

# If TRUE, there are NA values and we want to remove them from the timeseries
c11_cleaned <- c11[!is.na(c11$TotalPara), ]
```

```{r}
# Extract relevant columns into new dataframe
df <- c11_cleaned %>% select(Week, TotalFlies, TotalPara) # add %>% pull() if we want this to be a vector
head(df)


```

Initial visualisation of data to look for any basic patterns/trends
```{r}
# Plot the time series to visually inspect flies and parasites over time.
ggplot(df, aes(x = Week)) +
  geom_line(aes(y = TotalFlies), color = "blue", size = 1, linetype = "solid", alpha = 0.8) +
  geom_line(aes(y = TotalPara), color = "red", size = 1, linetype = "solid", alpha = 0.8) +
  scale_color_manual(values = c("Flies" = "blue", "Parasites" = "red"), name = "Legend Title") +
  labs(title = "Flies and Parasites Over Time",
       x = "Weeks",
       y = "Counts")

```

```{r}
# Convert dataframes into time series objects as this allows us to use more ts-specific functions
ts_flies <- ts(df$TotalFlies, start = 1, frequency = 1)
ts_para <- ts(df$TotalPara, start = 1, frequency = 1)

summary(ts_flies)
summary(ts_para)
```


```{r}
plot(ts_flies, main = "Plot of flies and Asobara parasite over time", xlab = "Week", ylab = "Count", col = "blue")
abline(h = mean(ts_flies), col = "blue", lty = 2)
lines(ts_para, col = "red")
abline(h = mean(ts_para), col = "red", lty = 2)
```

Can't use decomposition function because data is not long enough/lacks seasonality, which i guess should be expected? The functions are unable to detect a clear periodic pattern in the time series
```{r}
# Testing out some decomposition functions, but none of them seem to work. Just leaving them in here for future ref.

# decomposition <- decompose(ts_flies)
# decomposition_stl <- stl(ts_flies, s.window = "periodic")
# 
# library(seasonal)
# decomposition <- seas(ts_para)
# plot(decomposition)

```
Calculating simple moving averages
```{r}
# Moving averages. Smooth out short-term fluctuations to help highlight long-term trends.
library(TTR)
window_size <- 3 #Not sure how we choose window size tbh
MA_flies <- SMA(ts_flies, n = window_size)
MA_para <- SMA(ts_para, n = window_size)

# Plot the original time series
plot(ts_flies, main = "cage 11", col = "blue")

# Apply a simple moving average for smoothing
ma_smoothed <- stats::filter(ts_flies, rep(1/3, 3), sides = 2)

# Plot the smoothed time series
lines(ma_smoothed, col = "red")

# Add legend
legend("topright", legend = c("Original", "MA smoothed"), col = c("blue", "red"), lty = 1)
```

Checking for seasonality with the ADF test
```{r}
adf_result_flies <- adf.test(ts_flies)
print(adf_result_flies) # p-value > 0.05 indicates series is non-stationary

adf_result_para <- adf.test(ts_para)
print(adf_result_para) 
```

Seasonality can also be visually checked with ACF test
```{r}
acf_result_flies <- acf(ts_flies)
acf_result_para <- acf(ts_para)

```

Checking for correlations between host and parasite populations

```{r}
correlation_result <- cor(ts_flies, ts_para)
print(correlation_result)

# Scatter plot
plot(ts_flies, ts_para,
     pch = 16,
     xlab = "Host Population", ylab = "Parasite Population",
     main = "Scatter Plot of Host and Parasite Populations")

cross_correlation_result <- ccf(ts_flies, ts_para)

# Plot cross-correlation function
plot(cross_correlation_result, main = "Cross-Correlation Function")

```

Preparing timeseries data for forecasting
```{r}
# Split the timeseries in half, into training and testing data
n <- length(ts_flies)
splitpoint <- floor(n/2)
ts_flies_train <- ts_flies[1:splitpoint]
ts_flies_test <- ts(ts_flies[(splitpoint+1):n], start = splitpoint+1, frequency = 1)

ts_para_train <- ts_para[34:splitpoint]
ts_para_test <- ts(ts_para[(splitpoint+1):n], start = splitpoint+1, frequency = 1)
```

Naive Forecasting Method

The simplest forecasting method is to use the most recent observation as the forecast for the next observation. This is called a naive forecast and can be implemented using the 'naive()' function. This method may not be the best forecasting technique, but it often provides a useful benchmark for other, more advanced forecasting methods.
```{r}
naive_mod <- naive(ts_flies_test, h = 80)
summary(naive_mod)
print(naive_mod)
plot(naive_mod, main = "Fly population forecast from Naive method - cage 11", xlab = "Weeks", ylab = "Number of Flies")

```


Linear modelling
```{r}
# Linear model
time_index <- time(ts_flies)
lm_flies <- lm(ts_flies ~ time_index)
summary(lm_flies)
plot(lm_flies)
plot(residuals(lm_flies))

time_index <- time(ts_para)
lm_para <- lm(ts_para ~ time_index)
summary(lm_para)
plot(lm_para)

lm_fliespara <- lm(ts_flies ~ ts_para+time_index)
summary(lm_fliespara)

plot(lm_fliespara)
```


auto.ARIMA model
```{r}
# auto.ARIMA model - searches for best ARIMA model based on AIC
library(forecast)

autoarima_flies <- auto.arima(ts_flies)
autoarima_flies_forecast <- forecast(autoarima_flies, 52)
plot(fitted(autoarima_flies), col = "red", xlab = "Time", ylab = "Value", main = "auto.ARIMA Fitted Values Flies", xlim = c(1, length(ts_flies) + 52))
lines(autoarima_flies_forecast$mean, col = "pink")

plot(autoarima_flies_forecast, main = "ARIMA(3,1,3) forecast of flies - cage 11") # Shows historical timeseries, forecast for future points + uncertainty of those forecasts.

autoarima_para <- auto.arima(ts_para)
autoarima_para_forecast <- forecast(autoarima_para, 52)
plot(fitted(autoarima_para), col = "red", xlab = "Time", ylab = "Value", main = "auto.ARIMA Fitted Values Parasites", xlim = c(1, length(ts_flies) + 52))
lines(autoarima_para_forecast$mean, col = "pink")
# Predictions flatten out cos it's using the mean??

plot(autoarima_para_forecast, main = "ARIMA(1,1,1) forecast of parasites - cage 11")
```

Arima model and predictions
```{r}
# Fit arima model and generate predictions
library(forecast)
ts_flies_model <- auto.arima(ts_flies_train)
ts_flies_forecast <- forecast(ts_flies_model, h = length(ts_flies_test))

#Create new timeseries so that forecasted values begin from splitpoint and not 1 - this is needed for graphing later
forecast_time_vector <- seq(splitpoint+1, length.out = length(ts_flies_test), by = 1)
forecast_flies_values <- ts(ts_flies_forecast$fitted, start = splitpoint+1, frequency = 1)
print(forecast_flies_values)

# Plot the actual values and ARIMA forecast
plot(ts_flies_forecast)
lines(forecast_flies_values, col = "purple")
lines(ts_flies_test)
legend("topright", legend = c("Actual counts", "Forecasted counts"), col = c("black", "purple"), lty = 1)


plot(ts_flies_train, xlim = c(1, length(ts_flies)), ylim = c(-400, 400))
plot(ts_flies_test, xlim = c(1, length(ts_flies)), ylim = c(-400, 400))
par(new = TRUE)
plot(ts_flies_forecast$fitted, col = "orange")

```
```{r}
ts_para_model <- auto.arima(ts_para_train)
ts_para_forecast <- forecast(ts_para_model, h = length(ts_para_test))
print(ts_para_forecast)

#Create new timeseries so that forecasted values begin from splitpoint and not 1 - this is needed for graphing later
forecast_time_vector <- seq((splitpoint-34)+1, length.out = length(ts_para_test), by = 1)
print(forecast_time_vector)
forecast_para_values <- ts(ts_para_forecast$fitted, start = splitpoint+1, frequency = 1)
print(forecast_para_values)

# Plot the actual values and ARIMA forecast
plot(ts_para_forecast)
lines(forecast_para_values, col = "purple")
lines(ts_para_test)
legend("topright", legend = c("Actual counts", "Forecasted counts"), col = c("black", "purple"), lty = 1)

```

ARIMA model and predictions attempt 2
```{r}
library(stats)
stats_arima <- stats::arima(ts_flies_test, order=c(0,1,2))
print(stats_arima)
plot(stats_arima)
```


```{r}
library(Metrics)

#Mean Absolute Error (MAE), Mean Squared Error (MSE), and Root Mean Squared Error (RMSE).
mae_flies <- mae(ts_flies_prediction, ts_flies_test)
mse_flies <- mse(ts_flies_prediction$fitted, ts_flies_test)
rmse_flies <- sqrt(mse_flies)

cat(mae_flies, mse_flies, rmse_flies)
```

```{r}
plot(ts_flies, col = "blue", xlab = "Weeks", ylab = "Value", main = "ARIMA prediction vs actual")
lines(ts_flies_prediction$fitted, col = "red", lty = 2)

plot(ts_flies, col = "black", xlab = "Time", ylab = "Value", main = "ARIMA predictions vs Actual", xlim = c(1, length(ts_flies)))
lines(ts_flies_prediction$fitted, col = "pink")
lines(ts_flies_test)
```

```{r}
plot(ts_flies, xlim = c(1, length(ts_flies)), ylim =c(-400, 400), col = "black")
plot(ts_flies_prediction$fitted, xlim = c(1, length(ts_flies)), ylim = c(-400, 400), col = "pink")
par(new = TRUE)
plot(ts_flies_prediction, xlim = c(1, length(ts_flies)), ylim = c(-400, 400))

```

Calculating ARIMA values manually
```{r}
# Plot ACF
plot(ts_flies, main = "ts_flies", ylab = "Value")
acf_values = acf(ts_flies)


arimamodel = Arima(ts_flies, c(1,0,1)) #autoregressive, integrating(not important for us), moving average
str(arimamodel)

modelforecast = forecast(arimamodel, 48)
str(modelforecast)
modelforecast$mean

#Adding forecast to plot
plot(ts_flies)
lines(modelforecast$mean, col = "pink")
```

```{r}
# ADF test to check for stationarity
differencing_count <- 0

adf_test <- adf.test(ts_flies)
print(adf_test) # p-value > 0.05 indicates series is non-stationary
# write a while loop here to run differencing as long as series is non-stationary
while (adf_test$p.value > 0.05) {  # Adjust the significance level as needed
  ts_flies_diff <- diff(ts_flies)
  adf_test <- adf.test(ts_flies_diff)
  differencing_count <- differencing_count + 1
  print(adf_test)
}

cat("Number of differencing steps:", differencing_count, "\n")

```

```{r}
#Repeat d calculation for parasites
differencing_count <- 0
adf_test <- adf.test(ts_para)
print(adf_test) # p-value > 0.05 indicates series is non-stationary
# write a while loop here to run differencing as long as series is non-stationary
while (adf_test$p.value > 0.05) {  # Adjust the significance level as needed
  ts_para_diff <- diff(ts_para)
  adf_test <- adf.test(ts_para_diff)
  differencing_count <- differencing_count + 1
  print(adf_test)
}
cat("Number of differencing steps:", differencing_count, "\n")
```

d = differencing order - represents number of times the series needs to be differences for stationarity.

```{r}
# Calculating q - moving average order
print(ts_flies)
pacf_result <- pacf(ts_flies)
plot(pacf_result, main = "Partial Autocorrelation Function (PACF)")

```
The q value is the highest significant lag in the PACF plot - looks like 11 here?


```{r}
# ARIMA model
arima_flies <- arima(ts_flies, order = c(4, differencing_count, 11))
predicted_values <- predict(arima_flies, n.ahead = length(ts_flies))

# Plot the original data and the predicted values
plot(ts_flies, type = "l", col = "blue", xlab = "Time", ylab = "Value", main = "ARIMA vs Original Data", xlim = c(1, length(ts_flies) + 100))
lines(predicted_values$pred, col = "red", lty = 2)

# Add legend
legend("topright", legend = c("ts_flies", "ARIMA Predictions"), col = c("blue", "red"), lty = c(1, 2))
```


```{r}
# Basic plotting of time series all the other stuff we calculated
plot(ts_flies, xlab = "Time", ylab = "Value", main = "Time Series of Flies")
abline(lm_flies, col = "red")
# lines(MA_flies, col = "purple")
lines(fitted(arima_flies), col = "lightgreen")

plot(ts_para, xlab = "Time", ylab = "Value", main = "Time Series of Parasites")
abline(lm_para, col = "red")
# lines(MA_para, col = "purple")
lines(fitted(arima_para), col = "lightgreen")
```